Does It Scale? Static Output Approximation of PHP Web Applications Using Symbolic Execution

### 2017-01-01 paper.tex
22.12.2016
01.01.2016

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{paralist}

\usepackage{enumerate}
\usepackage{subcaption}
\usepackage{cleveref}
\usepackage{ucs}
\usepackage{framed}

@@

\lstset{
language=PHP,
basicstyle=\small\sffamily,
numbers=left,
numberstyle=\tiny,
frame=tb,
columns=fullflexible,
showstringspaces=false
}

@@

\section{Introduction}
With the emerge of the world wide web, dynamic web applications have become
widely popular. Various different implementation techniques have emerged,
ranging from technologies for programming languages (JSP, ASP .NET) and web
application frameworks for  script languages (Ruby On Rails, Django) to
languages tailored specifically to the domain of web applications. PHP \cite{phpNET} is
a programming language focused on server-side application development. As of 2012,
it was used by 78.8 percent of the ten million most popular websites (according
to Alexa popularity ranking) \cite{alexaPHP}, ranking 7th on the TIOBE
programming community index \cite{tiobePHP}, and was the ranked as the 6th most
popular language on GitHub \cite{githubPHP}.\\

One common property of all technologies for dynamic we b applications is the
\emph{staged computation} of output, or simplified: dynamic code generation.
Since based on both static values and runtime values, output code is generated step by step
dynamically, output can be dependent on application inputs. Dynamic web
applications incorporate various both server-side inputs (databases,
configuration) and client-side inputs respectively. Static
analysis of dynamic web applications, hence, has to consider different input
sources.

\emph{Symbolic execution} is a static analysis technique between program testing
and program proving. For a program, symbolic, i.e., abstract, yet fix inputs are
applied instead of concrete inputs \cite{Darringer1978,King1976}. The underlying
concept is to map program input (symbolically) to program output. Different classes of inputs may
result in variational output, representing dependencies of output and input.
Symbolic execution extends concrete execution semantics with semantics to
handle symbolic values. Hence, concrete execution semantics is contained in
symbolic execution semantics.\\

Symbolic execution has been used and studied for generating test suites, but
has more recently also been applied for approximating the output of web
applications. Knowledge about different output variants enabled useful analyses
for the domain of web applications, such as detecting and locating HTML
validation errors \cite{Nguyen:2011:AFH:2190078.2190142}. As for developers,
from an output model, outlines and callgraphs for easier IDE code navigation
\cite{Nguyen:2014:BCG:2635868.2635928}, or program slices
\cite{Nguyen:2015:CPS:2786805.2786872} can be computed. Existing work has been
integrated in the Varis plugin for the Eclipse IDE \cite{Nguyen:2015:VIS:2819009.2819140}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for smaller systems that are not
developed anymore. The previously mentioned tools though are only practical, if
for a given system the symbolic execution engine is scalable, i.e., will also
approximate reliable output for larger and more recent systems with good time
and space consumption.\\

To investigate the question, whether we can have practical tools built from the
symbolic execution engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such as
object-oriented programming, and extensive array support. We evaluated our
symbolic execution engine for large-scale and modern PHP systems.

During introduction of new semantics for object-orientation and advanced array
we encountered two main challenges. First, we walked a boundary between a
purely symbolic or mixed concrete and symbolic (concolic) execution engine
since concrete execution, e.g. foreach loops for arrays, increased code
coverage significantly. Second, we had to make a trade-off between the accuracy
of method calls and the scalability of this language feature as a method call
for different program state variants can have multiple targets.

Based on our observations and evaluation results, we learned that for dynamic
PHP constructs, expressions that are assembled dynamically, in combination with
symbolic values symbolic execution can be impractical. For object-orientation
and method calls, symbolic execution semantics can be a tightrope walk between
state space explosion and imprecision. Moreover, the nature of PHP challenges a
practical symbolic execution engine since for it to be accurate, a large number
of functions from the standard library needs to be supported.\\

Our key contributions in this paper include: (1) The tool infrastructure to
statically approximate the output for PHP web applications using symbolic
execution and (2) observations as well as an experience evaluation to identify
and measure PHP constructs impeding symbolic execution of PHP code.

\section{State Of the Art}

\section{Towards scalable symbolic execution}%

@@

\begin{table*}[t]
\centering 
	\begin{tabular}{p{4cm}rrrl}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{File Count} & \textbf{SLOC} &
	\textbf{Description}
	\\
	\midrule
	AddressBook & 8.2.5.2 & 239 & 51,907 & Adress Administration \\
	SchoolMate & 1.5.4 & 65 & 8,118 & School Administration \\
	TimeClock & 1.04 & 63 & 20,800 & XXX \\
	WebChess & 1.0.0 & 28 & 5,219 & XXX \\
	\midrule
	Drupal & 7.5.0 & 125 & 52,464 & Content Management System \\
	phpBB & 3.1.9 & 1,398 & 327,371 & Bulletin Board \\
	phpMyAdmin & 4.6.3 & 871 & 303,582 & Database Administration \\
	\midrule
	Anchor & 0.12.1 & 201 & 15,054 & Content Management System \\
	Kirby & XXX & XXX & XXX & Content Management System \\
	Automad & XXX & XXX & XXX & Content Management System \\
	Monstra & XXX & XXX & XXX & Content Management System \\
	Nibbleblog & XXX & XXX & XXX & Content Management System \\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

The coverage results are illustrated in Figure \ref{coverage}. We could replicate high coverage results for the four small-scale systems with a reach coverage and output coverage over 80 percent respectively. For the three modern/large-scale systems we measured poor coverage with reach coverage ranging from 5 to 30 percent and output coverage ranging from 4 to 20 percent. We measured medium to high coverage for the more recent small-scale systems: reach coverage ranging from 43 to 85 percent, output coverage ranging from 40 to 85 percent.
% RESULTS
% REACH COVERAGE and
% OUTPUT COVERAGE
\begin{figure}
	\centering	
	\resizebox{0.4\textwidth}{!}{%
		\input{diagrams/coverage}	
	}
	\caption{Coverage results for the PHP corpus}
	\label{coverage}
\end{figure}

@@

% CONTEXT DISTRIBUTION
% - FUNCTION CONTEXT
% - TEMPLATE CONTEXT
\begin{figure}
	\resizebox{0.4\textwidth}{!}{%
		\input{diagrams/context_distribution}
	}	
	\caption{Context distribution of never reached output candidates for the PHP
	corpus.}
	\label{fig:contexts}
\end{figure}

@@

% INCLUDE COVERAGE / SUCCESS RATE
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.5 \textwidth}    	
        \centering        
        % Include Coverage Plot        
		\input{diagrams/include_coverage}		
        \caption{Proportion of reached include expressions per project.}
        \label{fig:include_coverage_results1}
    \end{subfigure}%
	\hfill
    \begin{subfigure}[t]{0.5\textwidth}    
        \centering
        % Success Rate Plot        
        \input{diagrams/success_rate}        
        \caption{Rate of successful resolutions of reached include expressions
        per project.}
        \label{fig:include_coverage_results2}
    \end{subfigure}
    \label{fig:include_coverage_results}
    \caption{Include coverage and accuracy.}
\end{figure*}

@@

% Callback candidate description
\begin{figure}
	\centering
	\begin{subfigure}[t]{0.18\textwidth}
	\centering
		\includegraphics[scale=0.4]{images-paper/callgraph_a}
		\caption{Completely explained by callback candidates (no call from
		\texttt{main})}
		\label{fig:callgraph_a}
    \end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.25\textwidth}
	\centering
		\includegraphics[scale=0.4]{images-paper/callgraph_b}
		\caption{Partially (left; one call from \texttt{main}) or not at all  (right)
		explained by callback candidates}
		\label{fig:callgraph_b}
    \end{subfigure}
    \caption{Callgraph analysis: The bottom nodes represent functions containing missed output candidates, the top nodes depict entry points to them.}
    \label{fig:callgraph}
\end{figure}

@@

		\begin{lstlisting}[caption=A misplaced and unused \texttt{listing} with no real value to this paper.]
		function foo() {
   		   echo "foo";
		}
		\end{lstlisting}

\section{Related Work}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{paralist}
\usepackage{listings}
\usepackage{lipsum}
\usepackage{cite}
\usepackage{enumerate}
\usepackage{subcaption}
\usepackage{cleveref}
\usepackage{ucs}
\usepackage{framed}

@@

\lstset{
language=PHP,
basicstyle=\ttfamily\footnotesize,
basewidth={0.5em,0.4em}, % for other columns settings
%columns=fullflexible,
numbers=left,
numberstyle=\tiny,
frame=single,
showstringspaces=false,
xleftmargin=2em,
framexleftmargin=1.5em
}

@@

\section{Introduction}
With the emerging world wide web, dynamic web applications have become
widely popular. Various different implementation techniques have emerged,
ranging from technologies for programming languages (JSP, ASP .NET) and web
application frameworks for  script languages (Ruby On Rails, Django) to
languages tailored specifically to the domain of web applications. PHP \cite{phpNET} is
a programming language focused on server-side application development. As of 2012,
it was used by 78.8 percent of the ten million most popular websites (according
to Alexa popularity ranking) \cite{alexaPHP}, ranking 7th on the TIOBE
programming community index \cite{tiobePHP}, and was the ranked as the 6th most
popular language on GitHub \cite{githubPHP}.

One common property of all technologies for dynamic web applications is \emph{staged computation} of output: A dynamic web application as a whole consists of both static code, such as scripts, and dynamically generated code, such as responses to HTTP requests. The latter code, although assembled at runtime, may contain client-side parts of the web application, such as JavaScript. So, to study dynamic web applications in its entirety, we need to consider both static as well as dynamic aspects of systems.

%Since based on both static values and runtime values, output code is generated step by step
%dynamically, output can be dependent on application inputs. Dynamic web
%applications incorporate various both server-side inputs (databases,
%configuration) and client-side inputs respectively. Static
%analysis of dynamic web applications, hence, has to consider different input
%sources.

\emph{Symbolic execution} is a static analysis technique between program testing
and program proving. For a program, symbolic, i.e., abstract, yet fix inputs are
applied instead of concrete inputs \cite{Darringer1978,King1976}. The underlying
concept is to map program input (symbolically) to program output. Different classes of inputs may
result in variational output, representing dependencies of output and input.
As symbolic execution of a program returns an input-output-mapping for symbolic inputs, this analysis technique helps unfold the \emph{staged} nature of web applications: Since every feasible path is executed all possible output is contained in the symbolic output, and all dynamically generated variational parts of the web application can be analyzed.

%Symbolic execution extends concrete execution semantics with semantics to
%handle symbolic values. Hence, concrete execution semantics is contained in
%symbolic execution semantics. Symbolic execution has been used and studied for generating test suites, but
%has more recently also been applied for approximating the output of web
%applications.

Knowledge about different output variants enabled useful analyses
for the domain of web applications, such as detecting and locating HTML
validation errors \cite{Nguyen:2011:AFH:2190078.2190142}. As for developers,
from an output model, outlines and callgraphs for easier IDE code navigation
\cite{Nguyen:2014:BCG:2635868.2635928}, or program slices
\cite{Nguyen:2015:CPS:2786805.2786872} can be computed. Existing work has been
integrated in the Varis plugin for the Eclipse IDE \cite{Nguyen:2015:VIS:2819009.2819140}.

Despite the various use cases for analyses based on an approximated output
model, static output approximation for PHP web applications using symbolic
execution has so far only been evaluated for smaller systems that are not
maintained anymore. The previously mentioned tools though are only practical, if
for a given system the symbolic execution engine is scalable, i.e., will also
approximate output accurately for larger and more recent systems with good time
and space consumption.

To investigate the question, whether we can build practical tools from the
the symbolic execution engine, we re-implemented the engine with the specifications
of the previous symbolic execution semantics
\cite{Nguyen:2014:BCG:2635868.2635928} for PHP and additional features, such as
object-oriented programming, and extensive array support. We evaluated our
symbolic execution engine for large-scale and modern PHP systems.

During introduction of new semantics for object-orientation we encountered two main challenges. 
First, we walked a boundary between a purely symbolic or mixed concrete and symbolic execution engine
since concrete execution, e.g. foreach loops for arrays, increased code
coverage significantly. Second, we had to make a trade-off between the accuracy
of method calls and the scalability of this language feature as a method call
for different program state variants can have multiple targets.

Based on our observations and evaluation results, we learned that for dynamic
PHP constructs, expressions that are assembled dynamically, in combination with
symbolic values symbolic execution can be impractical. For object-orientation
and method calls, symbolic execution semantics can be a tightrope walk between
state space explosion and imprecision. Moreover, the nature of PHP challenges a
practical symbolic execution engine since for it to be accurate, a large number
of functions from the standard library needs to be supported.

Our key contributions in this paper include (1) a new tool infrastructure to statically approximate the output for PHP web applications using symbolic execution, and (2) a report of our observations with state-of-the-art web applications as well as an explanation on conceptual limitations for symbolic execution of PHP applications.

\section{State Of the Art}
In this section we present the motivation for building tools that address the \textcolor{blue}{development, quality and xxx} of dynamic web applications, but require an accurate static output approximation. All tools proposed so far \cite{Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140,Nguyen:2011:AFH:2190078.2190142} leverage a representation of all possible HTML client page output. Since any variant can be analyzed in particular, this enables tool support addressing the web application as a whole despite its \emph{staged} nature, where parts of the web application (client-side output) is just generated dynamically.

The work-flow for all these tools is illustrated in Figure \ref{fig:approach}, where prior to any analysis, the web application is symbolically executed and all client page output variants are parsed. To illustrate the concept of symbolic execution for this purpose consider the following example:

\begin{figure}[t!]
	\centering
	\includegraphics[width=0.45\textwidth]{images-paper/approach}

	\caption{Bener}
	\label{fig:approach}
\end{figure}

\begin{lstlisting}
<?php

echo 'Hello World!';
\end{lstlisting}

Based on this output representation, the so-called D-Model, subsequent analyses can be conducted:

\begin{compactitem}
\item \emph{PhpSync}: Using all variants of  HTML client page output, every single one can be statically checked for Markup Validity, i.e., if the web page conforms to syntactical specifications for HTML and other client-side languages. After validation errors are detected, auto-fixes are provided. Otherwise, the validation error can be traced back to source code responsible for the defect, which then can be refactored manually \cite{Nguyen:2011:AFH:2190078.2190142}.

\item \emph{WebSlice}: Bener bener
\item \emph{Varis}: Bener IDE bener
\end{compactitem}

\textcolor{blue}{\emph{D. Tool evaluation critique}
\begin{compactitem}
	\item Limitations/Assumptions
	\item Too small sample
\end{compactitem}
}

\section{Scalable Output Approximation}%

@@

\begin{table*}[t]
\centering 
	\begin{tabular}{lrp{4cm}rlrr}
	\toprule
	\textbf{System} & \textbf{Version} & \textbf{Classification} & \textbf{SLOC} & \textbf{\#files} & \textbf{\#OCs} & \textbf{\#includes}
	\\
	\midrule
	\textsc{AddressBook} & 8.2.5.2 & 239 & 51,907 & Adress Administration  & XXX & XXX\\
	\textsc{SchoolMate} & 1.5.4 & 65 & 8,118 & School Administration  & XXX & XXX\\
	\textsc{TimeClock} & 1.04 & 63 & 20,800 & XXX  & XXX & XXX\\
	\textsc{WebChess} & 1.0.0 & 28 & 5,219 & XXX  & XXX & XXX\\
	\midrule
	\textsc{Drupal} & 7.5.0 & 125 & 52,464 & Content Management System  & XXX & XXX\\
	\textsc{phpBB} & 3.1.9 & 1,398 & 327,371 & Bulletin Board  & XXX & XXX\\
	\textsc{phpMyAdmin} & 4.6.3 & 871 & 303,582 & Database Administration  & XXX & XXX\\
	\midrule
	\textsc{Anchor} & 0.12.1 & 201 & 15,054 & Content Management System  & XXX & XXX\\
	\textsc{Kirby} & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	\textsc{Automad} & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	\textsc{Monstra} & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	\textsc{Nibbleblog} & XXX & XXX & XXX & Content Management System  & XXX & XXX\\
	\bottomrule
	\end{tabular}
	\caption{Corpus of twelve PHP systems. The file count includes files with a .php,
	.inc, .bit or .module extension.}
	\label{corpus}
\end{table*}

@@

The coverage results are illustrated in Figure \ref{coverage}. We could replicate high coverage results for the four small-scale systems with a reach coverage and output coverage over 80 percent respectively. For the three modern/large-scale systems we measured poor coverage with reach coverage ranging from 5 to 30 percent and output coverage ranging from 4 to 20 percent. We measured medium to high coverage for the more recent small-scale systems: reach coverage ranging from 43 to 85 percent, output coverage ranging from 40 to 85 percent.
% RESULTS
% REACH COVERAGE and
% OUTPUT COVERAGE
\begin{figure}
	\centering	
%	\resizebox{0.4\textwidth}{!}{%
%		\input{diagrams/coverage}	
%	}
	\includegraphics[width=0.5\textwidth]{images-paper/coverage.eps}
	\caption{
		Results for output candidates: \emph{Coverage loss} repesents reached output candidates that were not part of the output. \emph{Simple file} and \emph{function definition} represent context classifications of missed output candidates.
	}
	\label{coverage}
\end{figure}

@@

% CONTEXT DISTRIBUTION
% - FUNCTION CONTEXT
% - TEMPLATE CONTEXT
%\begin{figure}
%	\resizebox{0.4\textwidth}{!}{%
%		\input{diagrams/context_distribution}
%	}	
%	\caption{Context distribution of never reached output candidates for the PHP
%	corpus.}
%	\label{fig:contexts}
%\end{figure}

@@

% INCLUDE COVERAGE / SUCCESS RATE
\begin{figure}
%    \centering
%    \begin{subfigure}[t]{0.5 \textwidth}    	
%        \centering        
%        % Include Coverage Plot        %
%		\input{diagrams/include_coverage}		
%        \caption{Proportion of reached include expressions per project.}
%        \label{fig:include_coverage_results1}
%    \end{subfigure}
%	\hfill
%    \begin{subfigure}[t]{0.5\textwidth}    
%        \centering
%        % Success Rate Plot        
%        \input{diagrams/success_rate}        
%        \caption{Rate of successful resolutions of reached include expressions
%        per project.}
%        \label{fig:include_coverage_results2}
%    \end{subfigure}

\includegraphics[width=0.5\textwidth]{images-paper/includes.eps}

    \label{fig:include_coverage_results}
    \caption{Include coverage and accuracy.}
\end{figure}

@@

% Callback candidate description
\begin{figure}[t!]
	\centering
	\includegraphics[scale=0.55]{images-paper/callgraph_cases}
    \caption{Callgraph analysis: The bottom nodes represent functions containing missed output candidates, the top nodes depict entry points to them.}
    \label{fig:callgraph}
\end{figure}

@@

\section{Related Work}
\begin{itemize}
	\item Hills et al. \cite{Hills:2013:ESP:2483760.2483786}
	\item Output approximation using context-free grammar construction; applications \cite{minamide2005static,wassermann2007sound}
\end{itemize}
